---
title: "Exploratory Data Analysis"
author: "Harlan Gillespie"
date: "29/08/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Dependencies

```{r Dependencies, message=FALSE}
library(SummarizedExperiment)
library(gplots)
library(DESeq2)
library(ggplot2)

```

## Plot colour palette

```{r}
tropical=  c('darkorange', 'dodgerblue', 'hotpink', 'limegreen', 'yellow')
palette(tropical)
par(pch=19)
```



## Load the dataset

Firstly, we can load the dataset obtained through the previous steps in the capstone project. These steps are outlined in the README.md file.

```{r cars}
colData <- DataFrame(read.delim("~/Coursera Capstone/PData.txt", stringsAsFactors=TRUE))
counts <- as.matrix(read.delim("~/Coursera Capstone/featureCount-data-ENTREZ.txt"))
rownames(counts) = counts[,1]
counts = counts[,-1]
```

Next, it is good practice to organise the dataset into a Summarized Experiment object.

```{r}
data.se = SummarizedExperiment(assays = list(counts = counts),  colData = colData)
```

Next, we can extract the count data matrix (edata) and the phenotype DataFrame (pdata)
```{r}
edata = assays(data.se)[[1]]
pdata = colData(data.se)
```

## Exploratory Data Analysis

We can show the sex and race of the samples using a table. There seems to be a bias towards african american (AA) samples over hispanic (HISP).

```{r}
table(pdata$Sex, pdata$Race)
```

Next we can look at the overall distribution of the dataset using boxplots.

```{r}
boxplot(edata, col = 2, range = 0)
```

It seems that outliers in the sample A1 that may make the rest of the dataset difficult to visualise. A data transformation is therefore necessary.


```{r}
summary(edata)
```

A1's larger variability, demonstrated by the boxplots and the summary table, may be due to its lower RNA integrity number. This will be something to monitor as the project continues.

Next, a log2 transformation can be tested and evaluated using the same boxplot method.

Removing genes that had little or no expression for all datasets is also a good practice.

```{r}
boxplot(log2(edata+1),col=2,range=0)
```

The data is now much more clear and easier to visualise.



```{r}
filt_edata = log2(edata[rowMeans(edata)>1,]+1)
boxplot(as.matrix(filt_edata),col=2)
```





```{r}
plot(filt_edata[,3], filt_edata[,4])
```


```{r}
heatmap(edata[rowMeans(edata)>500,], Rowv = NULL)
```

## Principal Component Analysis

```{r}
edata_centered = filt_edata - rowMeans(filt_edata)
svd1 = svd(edata_centered)
plot(svd1$d^2/sum(svd1$d^2),ylab="Percent Variance Explained",col=2)
```


```{r}
plot(svd1$v[,1],svd1$v[,2],ylab="2nd PC",xlab="1st PC", col=as.numeric(pdata$Group))
```

This PCA plot shows that the first PC forms two clusters predominantly explained by their phenotype (group).


```{r}
plot(svd1$v[,1],svd1$v[,2],ylab="2nd PC",xlab="1st PC", col=as.numeric(pdata$RIN))
```

This second PCA plot shows the second PC is loosely correlated with the RIN of each sample.


## Performing PCA with prcomp and ggplot

```{r}
count_pca = prcomp(filt_edata, scale = TRUE, center = TRUE)
dat = data.frame(X=count_pca$rotation[,1], Y=count_pca$rotation[,2], age_group = pdata$Group, RIN = pdata$RIN)
var_explained <- count_pca$sdev^2/sum(count_pca$sdev^2)
ggplot(dat, aes(x=X, y=Y, shape=age_group, color=RIN)) + geom_point(size=5) + labs(x=paste0("PC1: ",round(var_explained[1]*100,1),"%"),
       y=paste0("PC2: ",round(var_explained[2]*100,1),"%"))
```

